Ecosystem size and complexity are extrinsic drivers of food chain length in branching networks

### 2022-03-26 rmd/manuscript.Rmd
edit: manuscript

---
Title: "Ecosystem size and complexity are extrinsic drivers of food chain length in branching networks"
output:
  pdf_document:
    latex_engine: xelatex
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
bibliography: references.bib
header-includes:
  \usepackage{textgreek}
---

@@

We simulated the spatial distribution of disturbance intensity (i.e., proportional mortality in the event of stochastic regional disturbance) as follows. A disturbance value, $m_{h}$, was assigned randomly to upstream terminal patch $h$ as $logit~m_{h} \sim Normal(logit~\mu_m, \sigma_{h}^2)$, where $\mu_m$ is the average of disturbance intensity at headwaters in an ordinary scale. The disturbance values propagate downstream through a spatial autoregressive process as $logit~m_{down} = logit~m_{up}+ \epsilon$, where $m_{down}$ and $m_{up}$ are the disturbance values at down- and upstream patches, and the local environmental noise at the downstream patch $\epsilon$ was drawn from $Normal(0, \sigma_{l}^2)$. This autoregressive process resembles natural disturbance regime, where flood or drought disturbance cascades downstream with some local variation along the channel. At confluence patches, the disturbance value takes a weighted mean based on the relative size of the two contributing branches, $Q$ (the number of upstream contributing patches): $logit~m_{down} = \omega~\{logit~m_{1,up} +\epsilon_1\}+(1-\omega)~\{logit~m_{2,up} +\epsilon_2\}$, where $m_{1,up}$ and $m_{2,up}$ are the disturbance values of upstream contributing tributaries, $\epsilon_1$ and $\epsilon_2$ are independent local noises, and $\omega$ the relative size of one tributary over another defined as $\omega = \frac{Q_{1}}{Q_{1} + Q_{2}}$. Thus, the larger tributary has a greater influence on the downstream environment, as observed in natural river networks [@bendaNetworkDynamicsHypothesis2004]. Disturbance value distributions had medians close to $m$ with long tails (**Figure 1B**). The low-magnitude patches can be viewed as refugia, which supply individuals to impacted sites in subsequent time steps.

@@

```{=tex}
\begin{align}
K_{x} = K_{base}N_{p,up}^z
\end{align}
```
where $K_{base}$ is the carrying capacity at the upstream terminal patch, $N_{p,up}$ the number of patches upstream, and $z$ is the scaling exponent. We used $K_{base} = 100$ and $z = 0.54$ based on empirical relationships of gross primary production and watershed area [@finlayStreamSizeHuman2011].

@@

```{=tex}
\begin{align}
A_{i,x}(t+1) \sim Poisson(\bar{A}_{i,x}(t) + I_{i,x}(t) - E_{i,x}(t))
\end{align}
```
where $\bar{A}_{i,x}$ is the expected species abundance after within-patch trophic interactions, $I_{i,x}(t)$ the expected number of immigrants into patch $x$, and $E_{i,x}(t)$ the expected number of emigrants from patch $x$. The realized discrete number of individuals is drawn from a Poisson distribution to account for stochasticity in demographic and dispersal processes.

In each time step, internal dynamics occur first. The basal species is exposed to predation as well as other sources of mortality:

```{=tex}
\begin{align}
A_{B,x}'(t) = s_{0,B}\{1-z_{B,x}(t)\}A_{B,x}(t)
\end{align}
```
where $s_{0,i}$ is the background survival probability (density independent) and $z_{B,x}(t)$ the proportional mortality caused by predation. Subsequently, a density-dependent recruitment occurs:

```{=tex}
\begin{align}
\bar{A}_{B,x}(t) = \frac{r_{b}A_{B,x}'(t)}{1 + \frac{r_b - 1}{K_{x}}A_{B,x}'(t)}(1-\psi(t)~m_{x})
\end{align}
```
$r_b$ is the reproductive rate for survived individuals, $\psi(t)$ the random binary variable indicating the occurrence of regional disturbance (disturbance occurs if $\psi(t)=1$ otherwise $0$), and $m_x$ the proportional mortality caused by disturbance. $\psi(t)$ was drawn from a Bernoulli distribution with disturbance probability $p_m$. The predator-induced mortality $z_{B,x}(t)$ is determined as follows:

```{=tex}
\begin{equation}
z_{B,x}(t) = z_{BC,x}(t) + z_{BP,x}(t) =
\left\{
\begin{aligned}
&\frac{W_{BC,x}(t) + \delta_{x}(t) W_{BP,x}(t)}{A_{B,x}(t)} &&\text{if~$W_{BC,x}(t) + \delta_{x}(t) W_{BP,x}(t) < A_{B,x}(t)$}\\
&1 &&\text{if~$W_{BC,x}(t) + \delta_{x}(t) W_{BP,x}(t) > A_{B,x}(t)$}\\
\end{aligned}
\right.
\end{equation}
```
where $W_{ij,x}(t)$ is the total consumption of consumer $j$ on prey $i$ (number of individuals). We assumed the Holling Type-II functional response for predator-prey interactions:

```{=tex}
\begin{align}
W_{ij,x} = \frac{a_{ij}A_{i,x}A_{j,x}}{A_{j,x} + a_{ij}h_{ij}A_{i,x}}
\end{align}
```
$a_{ij}$ is the attack rate and $h_{ij}$ the handling time. The parameter $\delta_{x}(t)$ is the relative preference of intraguild predator $P$ to basal species $B$ over intraguild prey $C$. To be realistic, we assume that the preference parameter depends on the relative potential amount of energy obtained from each prey:

```{=tex}
\begin{align}
\delta_{x} = \frac{e_{BP} f_{BP}(A_{B,x})}{e_{BP} f_{BP}(A_{B,x}) + e_{CP} f_{CP}(A_{C,x})}
\end{align}
```
where $e_{ij}$ is the conversion efficiency from prey $i$ to consumer $j$, and $f_{ij}(A_{i,x}) = \frac{a_{ij}A_{i,x}}{1 + a_{ij}h_{ij}A_{i,x}}$. An analogous approach has been applied to a model with multiple prey choices [@shotanishijimaRolesAlternativePrey2014].

The intraguild prey $C$ is controlled by both prey availability and predation.

$$
\bar{A}_{C,x}(t) = e_{BC}s_{0,C}(1-z_{CP,x})z_{BC,x}A_{B,x}
$$

>>>>

---
title: "Ecosystem size and complexity are extrinsic drivers of food chain length in branching networks"
output:
  pdf_document:
    latex_engine: xelatex
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
bibliography: references.bib
header-includes:
  \usepackage{textgreek}
---

@@

We simulated the spatial distribution of disturbance intensity (i.e., proportional mortality in the event of stochastic regional disturbance) as follows. A disturbance value, $m_{h}$, was assigned randomly to upstream terminal patch $h$ as $logit~m_{h} \sim Normal(logit~\mu_m, \sigma_{h}^2)$, where $\mu_m$ is the average of disturbance intensity at headwaters in an ordinary scale. The disturbance values propagate downstream through a spatial autoregressive process as:

```{=tex}
\begin{align}
logit~m_{down} = logit~m_{up}+ \epsilon
\end{align}
```
where $m_{down}$ and $m_{up}$ are the disturbance values at down- and upstream patches, and the local environmental noise at the downstream patch $\epsilon$ was drawn from $Normal(0, \sigma_{l}^2)$. This autoregressive process resembles natural disturbance regime, where flood or drought disturbance cascades downstream with some local variation along the channel. At confluence patches, the disturbance value takes a weighted mean based on the relative size of the two contributing branches, $Q$ (the number of upstream contributing patches):

```{=tex}
\begin{align}
logit~m_{down} = \omega~(logit~m_{1,up} +\epsilon_1)+(1-\omega)~(logit~m_{2,up} +\epsilon_2)
\end{align}
```
$m_{1,up}$ and $m_{2,up}$ are the disturbance values of upstream contributing tributaries, $\epsilon_1$ and $\epsilon_2$ are independent local noises, and $\omega$ the relative size of one tributary over another defined as $\omega = \frac{Q_{1}}{Q_{1} + Q_{2}}$. Thus, the larger tributary has a greater influence on the downstream environment, as observed in natural river networks [@bendaNetworkDynamicsHypothesis2004]. Disturbance value distributions had medians close to $m$ with long tails (**Figure 1B**). The low-magnitude patches can be viewed as refugia, which supply individuals to impacted sites in subsequent time steps.

@@

```{=tex}
\begin{align}
K_{x} = K_{base}(N_{p,up} + 1)^z
\end{align}
```
where $K_{base}$ is the carrying capacity at the upstream terminal patch, $N_{p,up}$ the number of patches upstream, and $z$ is the scaling exponent. We used $K_{base} = 100$ and $z = 0.54$ based on an empirical relationship of gross primary production and watershed area [@finlayStreamSizeHuman2011].

@@

```{=tex}
\begin{align}
A_{i,x}(t+1) \sim Poisson(\bar{A}_{i,x}(t) + I_{i,x}(t) - E_{i,x}(t))
\end{align}
```
where $\bar{A}_{i,x}(t+1)$ is the expected species abundance after within-patch trophic interactions, $I_{i,x}(t)$ the expected number of immigrants into patch $x$, and $E_{i,x}(t)$ the expected number of emigrants from patch $x$. The realized discrete number of individuals is drawn from a Poisson distribution to account for stochasticity in demographic and dispersal processes.

***Internal dynamics.*** Each time step starts with internal dynamics. The basal species is exposed to predation as well as other sources of mortality:

```{=tex}
\begin{align}
A_{B,x}'(t) = s_{0,B}[1-z_{B,x}(t)]A_{B,x}(t)
\end{align}
```
where $s_{0,i}$ is the background survival probability (density independent) and $z_{B,x}(t)$ the proportional mortality caused by predation. Let $W_{B,x}(t)$ be the total number of basal species consumed by $C$ and $P$. The predator-induced mortality $z_{B,x}(t)$, the sum of proportional predation by $C$ ($z_{BC,x}(t)$) and $P$ ($z_{BP,x}(t)$), is then:

```{=tex}
\begin{equation}
z_{B,x}(t) = z_{BC,x}(t) + z_{BP,x}(t) =
\left\{
\begin{aligned}
&\frac{W_{B,x}(t)}{A_{B,x}(t)} &&\text{if~$W_{B,x}(t) < A_{B,x}(t)$}\\
&1 &&\text{if~$W_{B,x}(t) > A_{B,x}(t)$}\\
\end{aligned}
\right.
\end{equation}
```
and

```{=tex}
\begin{align}
W_{B,x}(t) = W_{BC,x}(t) + \delta_{x}(t) W_{BP,x}(t)
\end{align}
```
where $W_{ij,x}(t)$ is the population-wide consumption of consumer $j$ on prey $i$ (number of individuals), and the parameter $\delta_{x}(t)$ is the relative preference of intraguild predator to basal species over intraguild prey. We assume the Holling Type-II functional response for predator-prey interactions:

```{=tex}
\begin{align}
W_{ij,x}(t) = \frac{a_{ij}A_{i,x}(t)A_{j,x}(t)}{A_{j,x}(t) + a_{ij}h_{ij}A_{i,x}(t)}
\end{align}
```
$a_{ij}$ is the attack rate and $h_{ij}$ the handling time. To be realistic, the preference parameter is assumed to depend on the relative potential amount of energy obtained from each prey:

```{=tex}
\begin{align}
\delta_{x}(t) = \frac{e_{BP} f_{BP}(A_{B,x}(t))}{e_{BP} f_{BP}(A_{B,x}(t)) + e_{CP} f_{CP}(A_{C,x}(t))}
\end{align}
```
where $e_{ij}$ is the conversion efficiency from prey $i$ to consumer $j$, and $f_{ij}(A_{i,x}(t)) = \frac{a_{ij}A_{i,x}(t)}{1 + a_{ij}h_{ij}A_{i,x}(t)}$. An analogous approach has been applied to a model with multiple prey choices [@shotanishijimaRolesAlternativePrey2014]. Subsequently, a density-dependent recruitment occurs:

```{=tex}
\begin{align}
\hat{A}_{B,x}(t) = \frac{r_{b} A_{B,x}'(t)}{1 + \frac{r_b - 1}{K_{x}}A_{B,x}'(t)}
\end{align}
```
$r_b$ is the reproductive rate for survived individuals.

The population growth of the intraguild prey is controlled by conversion efficiency, number of prey captured, background mortality, and predation:

```{=tex}
\begin{equation}
\hat{A}_{C,x}(t) =
\left\{
\begin{aligned}
&s_{0,C}(1-z_{CP,x})e_{BC}W_{BC,x}(t) &&\text{if~$W_{B,x}(t) < A_{B,x}(t)$}\\
&s_{0,C}(1-z_{CP,x})\frac{e_{BC}W_{BC,x}(t)}{W_{B,x}(t)}A_{B,x} &&\text{if~$W_{B,x}(t) > A_{B,x}(t)$}\\
\end{aligned}
\right.
\end{equation}
```
Equation 11 assumes competition between $C$ and $P$ when their total consumption $W_{B,x}(t)$ exceeds the abundance of basal species $A_{B,x}(t)$, and the relative consumption $\frac{W_{BC,x}(t)}{W_{B,x}(t)}$ rests on $a_{ij}$, $h_{ij}$, and $A_{j,x}(t)$.

The population of the intraguild predator is fueled by two prey species:

```{=tex}
\begin{equation}
\hat{A}_{P,x}(t) =
\left\{
\begin{aligned}
&s_{0,P}\left[e_{BP}\delta_{x}(t)W_{BP,x}(t) + e_{CP}(1 - \delta_{x})W_{CP,x}(t)\right] &&\text{if~$W_{B,x}(t) < A_{B,x}(t)$}\\
&s_{0,P}\left[\frac{e_{BP}\delta_{x}(t)W_{BP,x}(t)}{W_{B,x}(t)}A_{B,x} + e_{CP}(1 - \delta_{x})W_{CP,x}(t)\right] &&\text{if~$W_{B,x}(t) > A_{B,x}(t)$}\\
\end{aligned}
\right.
\end{equation}
```
After recruitment, the constituent species experience stochastic regional disturbance, which causes proportional loss of $m$ to populations:

```{=tex}
\begin{align}
\bar{A}_{i,x}(t) = [1 - \psi(t)m_{x}]\hat{A}_{i,x}(t)
\end{align}
```
The binary random variable $\psi(t)$ expresses the occurrence of regional disturbance (disturbance occurs if $\psi(t)=1$ otherwise $0$), and $m_x$ is the disturbance intensity at patch $x$ as determined by the network generation process. $\psi(t)$ was drawn from a Bernoulli distribution with disturbance probability $p_m$.

***Dispersal.*** We simulated distance-dependent dispersal along a network. The number of emigrants from patch x at time t, $E_{i,x}(t)$, is estimated as the product of dispersal probability $p_{d}$ and species abundance $A_{i,x}$. The dispersal probability was assumed to be constant for all three species in each simulation run. Then, the number of emigrants from patch x is divided into immigrants, Iix(t) Ixy(t) to each patch y. The immigration probability at patch x and time t, ix(t) xy ξxy from patch x to y is calculated considering network structure and species dispersal capability as followsbased on the habitat structure among patches according to:

xy(t) =y,y≠xExy(t)e-θdxyxy,y≠xExy(t)e-θdxy

ix(t) = y,yxEiy(t)e-dxyxy,yxEiy(t)e-dxy

Where dxy is the distance between patch x and y, and θ controls the rate of decay. Then, the expected number of immigrants into patch x is calculated as Iix(t)=ix(t)xEix
